Preparing data from the Global Burden of Disease 2019 for METAHIT modelling

### 2021-11-08 metahit/gbd_process_2019.Rmd
Upload 2019 results for METAHIT

---
title:  "Preparing data from the Global Burden of Disease for METAHIT modelling"
author:
- "Jackson, Chris <chris.jackson@mrc-bsu.cam.ac.uk>"
date:   "`r Sys.time()`"
output:
    html_document:
        code_folding: show
params:
    date:   !r Sys.Date() - 1
editor_options: 
  chunk_output_type: console
---

@@

```{r}
library(tidyverse)
library(progress)
library(tempdisagg)
relative_path_gbd <- "~/scratch/chronic/metahit_data/2017"

@@

Read look-up table mapping between local authorities and city regions, and add names for regions that are not city regions.  No need to bother with matching regions to countries.

```{r,eval=FALSE}
regions <- c("East Midlands", "East of England", "Greater London", "North East England", 
                  'North West England', "South East England", "South West England", "Yorkshire and the Humber", "West Midlands")

# LTLAs to city regions and regions 
local_government_areas <- read_csv(file.path(relative_path_gbd, "mh_regions_lad_lookup.csv")) %>% rename(region="gordet")
lad <- 
    read.csv("~/work/chronic/belen/Local_Authority_District_to_Region__December_2017__Lookup_in_England_.csv")
utla <- 
  read.csv("~/work/chronic/belen/Lower_Tier_Local_Authority_to_Upper_Tier_Local_Authority__December_2017__Lookup_in_England_and_Wales.csv")

## Make UTLA to region lookup
lad_wales <- c("Isle of Anglesey", "Gwynedd", "Conwy", "Denbighshire", "Flintshire", 
"Wrexham", "Ceredigion", "Pembrokeshire", "Carmarthenshire", 
"Swansea", "Neath Port Talbot", "Bridgend", "Vale of Glamorgan", 
"Cardiff", "Rhondda Cynon Taf", "Caerphilly", "Blaenau Gwent", 
"Torfaen", "Monmouthshire", "Newport", "Powys", "Merthyr Tydfil")
utlareg <- utla %>% 
  filter(!duplicated(UTLA17NM)) %>%
  select(LTLA17NM, UTLA17NM) %>%
  rename("LAD17NM"=LTLA17NM) %>%
  left_join(lad, by="LAD17NM") %>%
  filter(!(LAD17NM %in% lad_wales))

Read Global Burden of Disease 2017 data

```{r,eval=FALSE}
gbd1 <- read_csv(file.path(relative_path_gbd, "IHME-GBD_2017_DATA-3e0b192d-1.csv"))

gbd2 <- read_csv(file.path(relative_path_gbd, "IHME-GBD_2017_DATA-3e0b192d-2.csv"))

gbd3 <- read_csv(file.path(relative_path_gbd, "IHME-GBD_2017_DATA-3e0b192d-3.csv"))

gbd4 <- read_csv(file.path(relative_path_gbd, "IHME-GBD_2017_DATA-3e0b192d-4.csv"))

gbd5 <- read_csv(file.path(relative_path_gbd, "IHME-GBD_2017_DATA-3e0b192d-5.csv"))

gbd <- rbind(gbd1, gbd2, gbd3, gbd4, gbd5)

## Attach geographical info to each GBD estimate
## Manual step - this should cover them all 
gbd$location_name[gbd$location_name=="St Helens"] <- "St. Helens"
gbd <- gbd %>% mutate(
  areatype = case_when(location_name %in% c("United Kingdom","Australia")  ~ "country",
                       location_name %in% c("England","Scotland","Wales","Northern Ireland")  ~ "nation",
                       location_name %in% regions  ~ "region",
                       location_name %in% unique(utla$LTLA17NM) ~ "ltla",
                       location_name %in% unique(utla$UTLA17NM) ~ "utla"
  )
)
stopifnot(all(!is.na(gbd$areatype)))

## Determine which estimates are in city regions 
gbd <- gbd %>%
  left_join(local_government_areas, by=c("location_name"="lad11nm")) %>%
  mutate(in_cityregion = !is.na(cityregion)) %>%
  filter(areatype %in% c("ltla","utla"))

## Next we want to decide whether to use LTLA/UTLA or region for areas not in city regions

@@

```{r,eval=FALSE}
disease_names_execute <- read_csv(file.path(relative_path_gbd, "disease_outcomes_lookup.csv")) %>%
    select(GBD_name, acronym) %>%
    mutate(disease = tolower(GBD_name))

DISEASE_SHORT_NAMES <- data.frame(disease = tolower(as.character(unique(gbd$cause_name))), 
                                  sname = tolower(abbreviate(unique(gbd$cause_name, max = 2))),
                                  stringsAsFactors = F) %>%
    mutate(is_not_dis = ifelse((str_detect(disease, "injuries") |
                                str_detect(disease, "All causes") |
                                str_detect(disease, "Lower respiratory infections")), 
                               1, 0) ) %>%
    mutate(is_not_dis = case_when(sname == "allc"  ~  2,
                                  sname == "lwri"  ~  1,
                                  ## Code for major depressive disorder (no deaths) and hypertensive heart disease (no incidence)
                                  sname == "hyhd"  ~  3,
                                  sname == "mjdd"  ~  3,
                                  TRUE  ~  is_not_dis)) %>%
    left_join(disease_names_execute, by="disease") %>%
    mutate(acronym = ifelse(str_detect(disease, "injuries"), disease, acronym),
           acronym = word(acronym, 1),
           males = ifelse(disease %in% c("uterine cancer", "breast cancer"), 0, 1),
           females = ifelse(disease %in% "prostate cancer", 0, 1),
           sname = gsub("'", '', sname),
           acronym = ifelse(is.na(acronym), "no_pif", acronym))
```

@@

```{r,eval=FALSE}
names(gbd) <- gsub(pattern = "_name", replacement = "", x = names(gbd))
gbd <- gbd %>%
    select(-contains("id")) %>%
    mutate(cause = tolower(cause))

saveRDS(gbd, file=file.path(relative_path_gbd, "GBD2017.rds"))

gbd <- readRDS(file=file.path(relative_path_gbd, "GBD2017.rds"))
```

@@

<!-- NOTE COUNTS FOR AREAS CONTAINING CITY REGIONS ARE NOT CONSTRAINED TO BE GREATER THAN TOTAL COUNTS FOR CITY REGIONS WITHIN THAT AREA. we will exclude non-city region data from this analysis.  see notes in noncityregions.Rmd -->

@@

Lower limits published as 0 and upper limits published as 1 are modified to be close but not equal to these bounds, while remaining consistent with the point estimate, since values of exactly 0 and 1 are inconsistent with the Beta distribution (TODO put this in disbayes::ci2num).

```{r,eval=FALSE}
gbdp <- 
    gbd %>%
    filter(metric == "Rate") %>% 
    filter(cause != "all causes") %>% 
    mutate(num = NA, denom = NA) %>%
    mutate(val = val/100000, lower = lower / 100000, upper = upper / 100000) %>%
    mutate(lower = if_else(lower==0, pmin(val/2, 0.00001), lower)) %>% 
    mutate(upper = if_else(upper==1, pmax((1+val)/2, 0.99999), upper))
```

The function `ci2num` in the `disbayes` package is then used to calculate the effective sample sizes.  This takes about 40 minutes on a fairly fast laptop.   It could be made a lot faster by using parallel processing.

```{r,eval=FALSE,cache=TRUE}
nest <- nrow(gbdp) 
pb <- progress_bar$new(total = nest) # show progress of computation
for (i in 1:nest){
  if (gbdp$val[i] < gbdp$upper[i] &
      gbdp$val[i] > gbdp$lower[i]) { 
    pb$tick()
    counts <- disbayes::ci2num(gbdp$val[i], gbdp$lower[i], gbdp$upper[i])
    gbdp$num[i] <- counts$num
    gbdp$denom[i] <- counts$denom
  }
}
saveRDS(gbdp, file=file.path(relative_path_gbd, "gbdp.rds"))
```

```{r,echo=FALSE}
gbdp <- readRDS(file=file.path(relative_path_gbd, "gbdp.rds"))
```

@@

```{r}
summary(gbdp$num/gbdp$denom - gbdp$val)
```

@@

```{r}
gbdp %>% select(val, num, denom) %>% arrange(desc(denom)) %>% head
```

@@

```{r,eval=FALSE}
gbdnum <- gbd %>%
  filter(metric=="Number") %>%
  select(measure, location, sex, age, cause, Number=val)
gbdp <- gbdp %>%
  left_join(gbdnum, by=c("measure","location","sex","age","cause")) %>%
  mutate(pop = Number / val)
saveRDS(gbdp, file=file.path(relative_path_gbd, "gbdp.rds"))
gbdp <- readRDS(file=file.path(relative_path_gbd, "gbdp.rds"))

@@

```{r}
gbdp <- gbdp %>%
  mutate(pop = if_else(is.na(pop), 5000, pop)) %>% 
  mutate(nodenom = is.na(denom) | (denom > pop),
         denom = if_else(is.na(denom), pop, denom),
         denom = if_else(denom > pop, pop, denom),
         num = ifelse(nodenom, round(val*denom), num))
```

@@

```{r} 
# Working with the data that has one row per five-year age group, 
# first construct 1-year counts as an extra column. 
gbdp <- gbdp %>%
  extract(age, c("from_age", "to_age"), "(.+) to (.+)", remove=FALSE, convert=TRUE) %>%
  mutate(from_age = case_when(age=="95 plus"  ~  95L,
                              age=="Under 5"  ~  0L,
                              TRUE  ~  from_age),
         to_age = case_when(age=="95 plus"  ~  99L,
                            age=="Under 5"  ~  4L,
                            TRUE  ~  to_age),
         agediff = to_age - from_age + 1,  # this will equal 5 
         num1yr = round(num/agediff),
         denom1yr = round(denom/agediff)) %>%
  rename(agegroup = age)

@@

# This takes about a minute 
gbdpyr <- group_modify(gbdp_grp, disagg) %>% 
  ungroup %>%
  left_join(gbdpyrd5, by=c("measure","location","ageyr","sex","cause",
                           "areatype", "cityregion", "region", "in_cityregion"))

@@

## Check for a few cases that the series match with the naive dividing by 5 method and the smooth methods
tloc <- sample(unique(gbd_disagg$location), 3)
test2 <- gbdpyrd5 %>%
  filter(cause=="ischemic heart disease", sex=="Male", measure=="Deaths",
         location %in% tloc)
gbdpyr %>% 
  filter(cause=="ischemic heart disease", sex=="Male", measure=="Deaths",
         location %in% tloc) %>%
  ggplot(aes(x=ageyr, y=denom, col=location, group=location)) + 
  geom_line() + geom_point() + 
  geom_line(data=test2, aes(y=denom1yr))

@@

## LTLAs from GBD aggregated to UTLAs
gbdpyrlagg  <- gbdpyr  %>%
  filter(!in_cityregion, areatype=="ltla") %>%
  left_join(utla, by=c("location"="LTLA17NM")) %>%
  group_by(measure, sex, ageyr, cause, UTLA17NM) %>%
  summarise(num=sum(num), denom=sum(denom)) %>%
  rename(location = "UTLA17NM") %>%
  mutate(areatype = "utla_agg") %>%
  ungroup %>%
  select(measure, sex, ageyr, cause, location, areatype, num, denom)

@@

## LTLAs and UTLAs from GBD aggregated to regions, excluding areas that are in city regions
gbdpyrreg  <- gbdpyr  %>%
  mutate(region = case_when(areatype=="ltla"  ~  lad$RGN17NM[match(location, lad$LAD17NM)],
                            areatype=="utla"  ~  utlareg$RGN17NM[match(location, utlareg$UTLA17NM)],
                            TRUE  ~  region
                              )) %>%
  filter(!in_cityregion) %>%
  group_by(measure, sex, ageyr, cause, region) %>%
  summarise(num=sum(num), denom=sum(denom)) %>%
  rename(location = "region") %>%
  mutate(areatype = "region") %>%
  ungroup %>%
  select(measure, sex, ageyr, cause, location, areatype, num, denom)

@@

10-year survival to use to inform remission rates, from the [ONS](https://www.ons.gov.uk/peoplepopulationandcommunity/healthandsocialcare/conditionsanddiseases/bulletins/cancersurvivalinengland/nationalestimatesforpatientsfollowedupto201).  These are not published by age.  They are aggregated over ages 15-99, whereas 1, 5 year survival is published by age group.

Methods [here](https://www.ons.gov.uk/peoplepopulationandcommunity/healthandsocialcare/conditionsanddiseases/methodologies/cancersurvivalstatisticalbulletinsqmi).   Assumes "conditional probabilities of surviving for patients diagnosed in current year are equal to those diagnosed over the full 10-year period of available data."  Caveats about trends through time.

Our approach allows uncertainty around this - could represent expected variations between age groups, given variations for 5 year survival.

Annual remission probability is computed from the 10 year survival probability.  Assumes that remission happened in one of the 10 years, with some unknown annual probability r such that the probability of remission in 10 years is equal to the 10 year survival probability s = 1 - (1-r)^10, so that r = 1 -(1-s)^(1/10).

Convert remission "rates" to numerators and denominators by assuming a denominator defined by the number of cancer cases over all ages and areas, but dividing all numerators and denominators by 100 to represent uncertainty about the true area-specific prevalence. For example, this gives a 95% credible interval for the annual breast cancer remission probability of 0.11 to 0.16.

```{r}
library(readxl)
rem <- read_excel("~/work/chronic/cancerdata/Figure_10__Predicted_10-year_net-survival_using_the_hybrid_approach_for_men_and_women_(aged_15_to_99_years).xls", range="A7:C26", na=":") %>%
  pivot_longer(cols=c("Men","Women"), names_to = "sex", values_to = "surv10") %>% 
  mutate(prem1 = 1 - (1 - surv10/100)^(1/10)) %>% 
  filter(Cancer %in% c("Breast","Uterus","Colorectal","Liver","Stomach","Lung"),
         !is.na(prem1)) %>%
  mutate(Cancer = fct_recode(Cancer, 
                             "Breast cancer"="Breast",
                             "Uterine cancer"="Uterus",
                             "Colon and rectum cancer"="Colorectal",
                             "Stomach cancer"="Stomach",
                             "Tracheal, bronchus, and lung cancer"="Lung"
                             #,
                             #"Liver cancer"="Liver"
                             ),
         gender = fct_recode(sex, "Male"="Men", "Female"="Women")) %>% 
  rename(disease = Cancer)

cancers <- c("Tracheal, bronchus, and lung cancer", "Colon and rectum cancer", "Stomach cancer", "Liver cancer", "Breast cancer", "Uterine cancer")

num_cancer <- gbddb %>% 
  filter(disease %in% cancers, 
         age %in% 15:99) %>% 
  group_by(gender, disease) %>% 
  summarise(ncases = sum(prev_num)) %>% 
  left_join(rem, by=c("disease","gender")) %>%
  mutate(rem_num = round(ncases*prem1 / 100),
         rem_denom = round(ncases / 100)) %>% 
  select(gender, disease, rem_num, rem_denom)
num_cancer

gbddb <- gbddb %>% 
  left_join(num_cancer) %>% 
  mutate(rem_num = replace_na(rem_num, 0),
         rem_denom = replace_na(rem_denom, 0))

>>>>

---
title:  "Preparing data from the Global Burden of Disease 2019 for METAHIT modelling"
author:
- "Jackson, Chris <chris.jackson@mrc-bsu.cam.ac.uk>"
date:   "`r Sys.time()`"
output:
    html_document:
        code_folding: show
params:
    date:   !r Sys.Date() - 1
editor_options: 
  chunk_output_type: console
---

@@

```{r}
library(tidyverse)
library(progress)
library(disbayes) ## Install
relative_path_execute <- "~/scratch/chronic/mh-execute"
relative_path_gbd <- "~/scratch/chronic/mh-mslt/input/gbd/GBD2019/METAHIT/"
```

Read look-up table mapping between local authorities and city regions, and add names for regions that are not city regions

```{r,eval=FALSE}
names_non_cr <- c("United Kingdom", "England", "East Midlands", "East of England", "Greater London", "North East England", 
                  'North West England', "South East England", "South West England", "West Midlands", "Yorkshire and the Humber", 
                  "Northern Ireland", "Scotland", "Wales")
local_government_areas <- read_csv(file.path(relative_path_execute, "inputs/mh_regions_lad_lookup.csv"), show_col_types = FALSE) %>% 
  rename(region="gordet")
for (i in names_non_cr){
  local_government_areas <- rbind(local_government_areas, rep(i, ncol(local_government_areas)))
}
local_government_areas  <- local_government_areas %>%
#    dplyr::filter(!is.na(cityregion)) %>%
    dplyr::rename(location = lad11nm) %>%
    dplyr::mutate(location = gsub('St. Helens', 'St Helens', location))

regions <- c("East Midlands", "East of England", "Greater London", "North East England", 
                  'North West England', "South East England", "South West England", "Yorkshire and the Humber", "West Midlands")
lad <- 
    read.csv("~/work/chronic/disbayes/metahit/geodata/Local_Authority_District_to_Region_(December_2018)_Lookup_in_England.csv")
utla <- 
  read.csv("~/work/chronic/disbayes/metahit/geodata/Lower_Tier_Local_Authority_to_Upper_Tier_Local_Authority_(December_2018)_Lookup_in_England_and_Wales_v2.csv")

## Make UTLA to region lookup
lad_wales <- c("Isle of Anglesey", "Gwynedd", "Conwy", "Denbighshire", "Flintshire", 
"Wrexham", "Ceredigion", "Pembrokeshire", "Carmarthenshire", 
"Swansea", "Neath Port Talbot", "Bridgend", "Vale of Glamorgan", 
"Cardiff", "Rhondda Cynon Taf", "Caerphilly", "Blaenau Gwent", 
"Torfaen", "Monmouthshire", "Newport", "Powys", "Merthyr Tydfil")
utlareg <- utla %>% 
  filter(!duplicated(UTLA18NM)) %>%
  select(LTLA18NM, UTLA18NM) %>%
  rename("LAD18NM"=LTLA18NM) %>%
  left_join(lad, by="LAD18NM") %>%
  filter(!(LAD18NM %in% lad_wales))

@@

Read Global Burden of Disease data

I'm doing the geographical manipulations bit differently here, to obtain a mutually exclusive set of LTLAs and UTLAs that cover the whole of England.   We can aggregate some of those to get the city regions, and aggregate the remaining areas by region.

GBD look to be using 2018 area codings in their 2019 data, e.g. Bournemouth and Poole are separate in 2019, but April 2019 created new unitary authority "Bournemouth, Christchurch and Poole".

```{r,eval=FALSE}
list_of_files <- list.files(path = relative_path_gbd,
                            pattern = "\\.csv$",
                            full.names = TRUE)
gbd <- do.call(rbind, lapply(list_of_files, read.csv))

## Attach geographical info to each GBD estimate
## Manual step - this should cover them all 
gbd$location_name[gbd$location_name=="St Helens"] <- "St. Helens"
gbd <- gbd %>% mutate(
  areatype = case_when(location_name %in% c("United Kingdom","Australia")  ~ "country",
                       location_name %in% c("England","Scotland","Wales","Northern Ireland")  ~ "nation",
                       location_name %in% regions  ~ "region",
                       location_name %in% unique(utla$LTLA18NM) ~ "ltla",
                       location_name %in% unique(utla$UTLA18NM) ~ "utla"
  )
)
stopifnot(all(!is.na(gbd$areatype)))

## Determine which estimates are in city regions 
gbd <- gbd %>%
  left_join(local_government_areas, by=c("location_name"="location")) %>%
  mutate(in_cityregion = !is.na(cityregion)) %>%
  filter(areatype %in% c("ltla","utla"))

## Check that deaths add up to 500,000, and illustrate number of deaths by cause
egd <- c("All causes", "Ischemic heart disease", "Multiple myeloma")
gbd %>% 
  filter(measure_name=="Deaths", 
         metric_name=="Number" 
         ) %>% 
  group_by(cause_name) %>%
  select(cause_name, val) %>%
  summarise(num = sum(val)) %>%
  arrange(desc(num)) %>%
  as.data.frame()
```

@@

```{r,eval=FALSE}
disease_names_execute <- read_csv(file.path(relative_path_execute, "inputs/dose_response/disease_outcomes_lookup_new.csv")) %>% 
  select(GBD_name, acronym) %>%
  mutate(disease = tolower(GBD_name))

DISEASE_SHORT_NAMES <- disease_names_execute %>%
  mutate(sname = abbreviate(disease)) %>%
  mutate(is_not_dis = ifelse((grepl("injuries", disease) | # grepl determines if the disease string contains 'injuries'
                                grepl("all causes", disease) |
                                grepl("lower respiratory infections", disease)), 
                             1, 0)) %>%
  mutate(is_not_dis = case_when(sname == "allc"  ~  2,
                                sname == "lwri"  ~  1,
                                sname == "npls" ~ 2,
                                sname == "crdd" ~ 2,
                                TRUE  ~  is_not_dis)) %>%
  mutate(
    males = ifelse(disease %in% c("uterine cancer", "breast cancer"), 0, 1),
    females = ifelse(disease %in% "prostate cancer", 0, 1),
    sname = gsub("'", '', sname),
    acronym = ifelse(is.na(acronym), sapply(strsplit(disease, " "), head, 1), acronym))
saveRDS(DISEASE_SHORT_NAMES, paste0(relative_path_gbd, "/DISEASE_SHORT_NAMES.rds"))
```

@@

```{r,eval=FALSE}
names(gbd) <- gsub(pattern = "_name", replacement = "", x = names(gbd))
gbd_temp <- gbd %>%
    dplyr::select(-contains("id")) %>%
    dplyr::mutate(cause = tolower(cause)) %>%
    pivot_wider(names_from="metric", values_from=c("val", "upper", "lower")) %>%
    mutate(pop = val_Number*100000/val_Rate) %>%
    pivot_longer(cols = val_Percent:lower_Number,
                 names_to = "metric",
                 values_to = "val") %>%
    separate(metric, c("metric2", "metric")) %>%
    pivot_wider(names_from = "metric2", values_from = "val")
save(gbd_temp, local_government_areas, DISEASE_SHORT_NAMES, 
     file=file.path(relative_path_gbd, "/GBD2019.rda"))

load(file=file.path(relative_path_gbd, "/GBD2019.rda"))
```

@@

@@

Lower limits published as 0 and upper limits published as 1 are modified to be close but not equal to these bounds, while remaining consistent with the point estimate, since values of exactly 0 and 1 are inconsistent with the Beta distribution

```{r,eval=FALSE}
gbdp <- 
    gbd_temp %>%
    dplyr::filter(metric == "Rate") %>% 
    dplyr::filter(!cause %in% c("all causes", "road injuries", "pedestrian road injuries", "cyclist road     injuries", "motorcyclist road injuries", "motor vechicle road injuries", "other road injuries", "motor vehicle road injuries", "depressive disorders" , "hypertensive heart disease")) %>%
    dplyr::mutate(val = val/100000, lower = lower/100000, upper = upper/100000) %>%
    dplyr::mutate(lower = if_else(lower<=0, pmin(val/2, 0.00001), lower)) %>% 
    dplyr::mutate(upper = if_else(upper>=1, pmax((1+val)/2, 0.99999), upper))
# CJ Breaks at i = 27954 since upper credible limit is greater than 1. 
# There are 29 rows with upper > 1 
# Lower respiratory infections, age 95. 
# Perhaps from high chance of more than one event in a year. 
# Restrict credible limits on indiv probs in these cases to be less than 1
# Or could have used actual population size 
```

The function `ci2num` in the `disbayes` package is then used to calculate the effective sample sizes.  This is computationally intensive.

```{r,eval=FALSE,cache=TRUE}

library(foreach)
library(doParallel)
registerDoParallel(3)

nest <- nrow(gbdp)
numdenom <- foreach (i=1:nest, .combine=rbind) %dopar% {
  if (gbdp$val[i] < gbdp$upper[i] &
      gbdp$val[i] > gbdp$lower[i]) { 
    counts <- disbayes:::ci2num(gbdp$val[i], gbdp$lower[i], gbdp$upper[i], denom0=gbdp$pop[i])
    c(rowid = i, num=counts$num, denom=counts$denom)
  }
}
numdenom <- as.data.frame(numdenom)
rownames(numdenom) <- NULL
saveRDS(numdenom, file=file.path(relative_path_gbd, "numdenom.rds"))

gbdp <- gbdp %>% 
  mutate(rowid = row_number()) %>%
  left_join(numdenom, by="rowid") %>% 
  mutate(rowid= NULL)

saveRDS(gbdp, file=file.path(relative_path_gbd, "GBD2019.rds"))
```

```{r,echo=FALSE}
gbdp <- readRDS(file=file.path(relative_path_gbd, "GBD2019.rds"))
```

@@

```{r}
summary(gbdp$num/gbdp$denom - gbdp$val, na.rm=TRUE)
```

@@

```{r}
gbdp %>% dplyr::select(val, num, denom) %>% dplyr::arrange(desc(denom)) %>% head
```

@@

```{r,eval=FALSE}
gbdnum <- gbd_temp %>%
  dplyr::filter(metric=="Number") %>%
  dplyr::select(measure, location, sex, age, cause, Number=val)
gbdp <- gbdp %>%
  left_join(gbdnum, by=c("measure","location","sex","age","cause")) %>%
  dplyr::mutate(pop = Number / val)
```

@@

Use an arbitrary plausible number (5000) where the population size was indeterminate

```{r}
gbdp <- gbdp %>%
  mutate(pop = if_else(is.na(pop), 5000, pop)) %>% 
  dplyr::mutate(nodenom = is.na(denom) | (denom > pop),
         denom = if_else(is.na(denom), pop, denom),
         denom = if_else(denom > pop, pop, denom),
         num = ifelse(nodenom, round(val*denom), num))
summary(gbdp$num/gbdp$denom - gbdp$val, na.rm=TRUE)
```

That looks better.

@@

```{r} 
# Working with the data that has one row per five-year age group, 
# first construct 1-year counts as an extra column. 
gbdp <- gbdp %>%
  tidyr::extract(age, c("from_age", "to_age"), "(.+) to (.+)", remove=FALSE, convert=TRUE) %>%
  mutate(from_age = case_when(age=="95 plus"  ~  95L,
                              age=="Under 5"  ~  0L,
                              TRUE  ~  from_age),
         to_age = case_when(age=="95 plus"  ~  99L,
                            age=="Under 5"  ~  4L,
                            TRUE  ~  to_age),
         agediff = to_age - from_age + 1,  # this will equal 5 
         num1yr = round(num/agediff),
         denom1yr = round(denom/agediff)) %>%
  rename(agegroup = age)

@@

# This takes about a minute 
library(tempdisagg)
gbdpyr <- group_modify(gbdp_grp, disagg) %>% 
  ungroup %>%
  left_join(gbdpyrd5, by=c("measure","location","ageyr","sex","cause",
                           "areatype", "cityregion", "region", "in_cityregion"))

@@

## Check for a few cases that the series match with the naive dividing by 5 method and the smooth methods
tloc <- sample(unique(gbdpyr$location), 3)
test2 <- gbdpyrd5 %>%
  filter(cause=="ischemic heart disease", sex=="Male", measure=="Deaths",
         location %in% tloc)
gbdpyr %>% 
  filter(cause=="ischemic heart disease", sex=="Male", measure=="Deaths",
         location %in% tloc) %>%
  ggplot(aes(x=ageyr, y=denom, col=location, group=location)) + 
  geom_line() + geom_point() + 
  geom_line(data=test2, aes(y=denom1yr))

@@

## LTLAs from GBD aggregated to UTLAs
gbdpyrlagg  <- gbdpyr  %>%
  filter(!in_cityregion, areatype=="ltla") %>%
  left_join(utla, by=c("location"="LTLA18NM")) %>%
  group_by(measure, sex, ageyr, cause, UTLA18NM) %>%
  summarise(num=sum(num), denom=sum(denom)) %>%
  rename(location = "UTLA18NM") %>%
  mutate(areatype = "utla_agg") %>%
  ungroup %>%
  select(measure, sex, ageyr, cause, location, areatype, num, denom)

@@

## LTLAs and UTLAs from GBD aggregated to regions, excluding areas that are in city regions
gbdpyrreg  <- gbdpyr  %>%
  mutate(region = case_when(areatype=="ltla"  ~  lad$RGN18NM[match(location, lad$LAD18NM)],
                            areatype=="utla"  ~  utlareg$RGN18NM[match(location, utlareg$UTLA18NM)],
                            TRUE  ~  region
                              )) %>%
  filter(!in_cityregion) %>%
  group_by(measure, sex, ageyr, cause, region) %>%
  summarise(num=sum(num), denom=sum(denom)) %>%
  rename(location = "region") %>%
  mutate(areatype = "region") %>%
  ungroup %>%
  select(measure, sex, ageyr, cause, location, areatype, num, denom)

@@

10-year survival to use to inform remission rates, from the [ONS](https://www.ons.gov.uk/peoplepopulationandcommunity/healthandsocialcare/conditionsanddiseases/bulletins/cancersurvivalinengland/nationalestimatesforpatientsfollowedupto2017

https://www.ons.gov.uk/peoplepopulationandcommunity/healthandsocialcare/conditionsanddiseases/datasets/cancersurvivalratescancersurvivalinenglandadultsdiagnosed

1-p is prob of dying in 10 years 
1-r is prob of not remitting ie prob of stiking w cancer or dying in 1 year 
(1-r)^10 is prob of not remitting in 10 years 
p10 = 1 - (1-r)^10 is prob of remitting in 10 years 
1 - (1-p10)  ^ (1/10)

```{r}
cancers_keep <- c("Breast","Colorectal","Larynx","Liver","Lung","Myeloma","Stomach","Uterus")
rem <- readxl::read_excel("~/work/chronic/cancerdata/adultcancersurvivaltables.xlsx", 
                  sheet="4. Predicted Survival", 
                  skip=5,
                  na = ":") %>%
  select("Cancer", "Sex", standardised="Age group",  agegroup = "...4", 
         surv10="10-year \r\nsurvival (%)",
         surv10lower="Lower 95% CI...13", surv10upper="Upper 95% CI...14") %>%
  mutate(surv10 = 1 - (1 - surv10/100)^(1/10),
         surv10lower = 1 - (1 - surv10lower/100)^(1/10),
         surv10upper = 1 - (1 - surv10upper/100)^(1/10)
         ) %>% 
  fill(Cancer, Sex, standardised) %>%
  filter(standardised== "Non-standardised", 
         Sex != "Persons",
         agegroup != "All ages",
         Cancer %in% cancers_keep,
         !is.na(surv10)) %>%
  tidyr::extract(agegroup, c("agefrom", "ageto"), "([[:digit:]]+)-([[:digit:]]+)", convert=TRUE) %>%
  mutate(agediff = ageto - agefrom)

Convert these to numerators and denominators, using ci2num

```{r}
numdenom <- matrix(nrow=nrow(rem), ncol=2)
for (i in 1:nrow(rem)){
  numdenom[i,] <-  unlist(disbayes::ci2num(rem$surv10[i], rem$surv10lower[i], rem$surv10upper[i]))
}
numdenom <- as.data.frame(numdenom)
names(numdenom) <- c("rem_num","rem_denom")
rem <- cbind(rem, numdenom)
```

Disaggregate into single years of age.   Can't work out how to use tempdisagg to get smooth disaggregations for this, since the age intervals are uneven.    Just put equal numbers into each year of age for the moment.

```{r}
disagg <- function(dat, key){ 
  res <- with(dat, { 
    data.frame( 
      ageyr = rep(agefrom,agediff) + sequence(agediff) - 1,
      num = predict(td(rem_num ~ 1, to=5, method="fast")),
      denom = predict(td(rem_denom ~ 1, to=5, method="fast"))
    ) } )
  res 
}
#dat1 <- rem[rem$Cancer=="Colorectal" & rem$Sex=="Men",]
#td(dat1$rem_num ~ 0 + dat1$agediff, to=10, method = "fast")

inds <- rep(1:nrow(rem), rem$agediff+1)
remyr <- rem[inds,] %>%
  mutate(ageyr = agefrom + sequence(rem$agediff+1) - 1, 
           rem_num = round(rem_num / (agediff + 1)),
         rem_denom = round(rem_denom / (agediff + 1))) %>%
  select(Cancer, Sex, ageyr, rem_num, rem_denom)

## Childhood cancer survival in England only published for all cancers combined 
## Repeat the 15-year values.  Should also inflate the uncertainty

remyr15 <- remyr %>% filter(ageyr==15)
inds <- rep(1:nrow(remyr15), each=15)
remyr014 <- remyr15[inds,] %>% 
  mutate(ageyr = rep(0:14, nrow(remyr15)),
         rem_num = round(rem_num/2),
         rem_denom = round(rem_denom/2))

remyr <- remyr %>% 
  full_join(remyr014) %>%
  arrange(Cancer, Sex, ageyr) %>%
  mutate(Cancer = fct_recode(Cancer, 
                             "Breast cancer"="Breast",
                             "Uterine cancer"="Uterus",
                             "Colon and rectum cancer"="Colorectal",
                             "Stomach cancer"="Stomach",
                             "Tracheal, bronchus, and lung cancer"="Lung",
                             "Liver cancer"="Liver",
                             "Head and neck cancer"="Larynx", 
                             "Multiple myeloma"="Myeloma"
                             ),
         gender = fct_recode(Sex, "Male"="Men", "Female"="Women")) %>%
  select(-Sex) %>%
  rename(disease = Cancer, age=ageyr)

# Merge the head and neck cancers, assume larynx remission applies to all

hncancers <- c("Larynx cancer", "Lip and oral cavity cancer", "Nasopharynx cancer", "Other pharynx cancer")
gbd_hnc <- gbddb %>% 
  filter(disease %in% hncancers) %>% 
  group_by(age, gender, area) %>%
  summarise(mort_num=sum(mort_num), inc_num=sum(inc_num), 
            prev_num=sum(prev_num), 
            mort_denom=sum(mort_denom), inc_denom=sum(inc_denom), 
            prev_denom=sum(prev_denom),
            .groups = "drop") %>%
  mutate(disease = "Head and neck cancer")
gbddb <- gbddb %>%
  filter(!(disease %in% hncancers)) %>% 
  full_join(gbd_hnc)

gbddb <- gbddb %>% 
  left_join(remyr, by=c("age","gender","disease")) %>% 
  mutate(rem_num = replace_na(rem_num, 0),
         rem_denom = replace_na(rem_denom, 0))

saveRDS(gbddb, file=file.path(relative_path_gbd, "gbddb.rds"))
gbddb <- readRDS(file=file.path(relative_path_gbd, "gbddb.rds"))
saveRDS(gbddb, file=file.path("~/work/chronic/disbayes/metahit", "gbddb.rds"))

gbddb <- gbddb %>% 
  filter(!(disease %in% c("Chronic myeloid leukemia", # no deaths 
                         "Lower respiratory infections", # not chronic 
                         "Cyclist road injuries"     # not chronic 
                         )))
